Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.

### 2018-03-06 manuscript/manuscript.tex
first pass at materials and methods

We can clearly observe how shifting amino-acid preferences degrade the accuracy of the ExpCM by fitting them to containing increasingly diverged sequences.
Specifically, for H1 and H3 HAs, we created three phylogenetic trees: a ``low'' divergence tree that contains only sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
\ref{fig:compete} shows the subtrees containing each of these sets of HA sequences.
For each subtree, we examined the congruence between site-specific natural selection and the amino-acid preferences measured in the deep mutational scanning experiment. 
We used the ExpCM stringency parameter $\beta$ to measure this congruency~(\ref{eq:Frxy}). 
In essence, $\beta$ rescales the deep mutational scanning measurements~(\ref{fig:compete}). 
Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with a greater stringency, indicating a strong congruence between natural selection and the experimental preferences.

@@

\newpage
\section*{Materials and Methods}

\subsection*{Influenza hemagglutinin sequences.}
We downloaded all full-length protein-coding sequences for each of the 18 influenza HA subtypes. 
We used \texttt{phydms\_prepalignment} \citep{hilton2017phydms} to filter and align the sequences. 
We subsampled to five per year plus the required sequences. 
Total of 87 sequences. 
Inferred tree topology using \texttt{RAxML}.

\subsection*{Simulations.}
How was the model fit? 
How many simulations? 
How was the tree constructed?

\subsection*{Expected Asymptotic Sequence Identity.}
How were the models fit? 
What is the equation? 
What is the equation for the $\omega_r$ term? (does it include the ExpCM $\omega$?)
How is the GY94$\omega_r$ model constructed? ie How does it interact with the $\omega$ value in the GY94 model?

\subsection*{Deep mutational scanning amino acid preferences.}
Where did we get the preferences? 
How were the average preferences calculated? 
How was the ``hybrid" sequence made? 
Numbering scheme.

\subsection*{Substitution models}
\subsubsection*{GY94 models}
\subsubsection*{ExpCMs}
We recap the \textbf{Exp}erimentally Informed \textbf{C}odon \textbf{M}odel (ExpCM) \citep{bloom2014experimentally,bloom2014informed,bloom2017identification,hilton2017phydms} to introduce nomenclature.

In an ExpCM, rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is written in mutation-selection form~\citep{halpern1998evolutionary,mccandlish2014modeling,spielman2015relationship} as
\begin{equation}
P_{r,xy} = Q_{xy} \times F_{r,xy}
\end{equation}
where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, and $F_{r,xy}$ is proportional to the probability that this mutation fixes.
The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form:
\begin{equation}
Q_{xy} = 
\begin{cases}
\phi_w & \mbox{if $x$ and $y$ differ by a transversion to nucleotide $w$} \\
\kappa \phi_w & \mbox{if $x$ and $y$ differ by a transition to nucleotide $w$} \\
0 & \mbox{if $x$ and $y$ differ by $>1$ nucleotide.}
\end{cases}
\end{equation}
The $\kappa$ parameter represents the transition-transversion ratio, and the $\phi_w$ values give the expected frequency of nucleotide $w$ in the absence of selection on amino-acid substitutions, and are constrained by $1 = \sum_w \phi_w$.

@@

The ExpCM stationary state frequency $p_{r,x}$ of codon $x$ at site $r$ is~\citep{bloom2017identification} 
\begin{equation}
\label{eq:p_rx}
p_{r,x} = \frac{\left(\pi_{r,\mathcal{A}\left(x\right)}\right)^{\beta} \phi_{x_0} \phi_{x_1} \phi_{x_2}}{\sum_z \left(\pi_{r,\mathcal{A}\left(z\right)}\right)^{\beta} \phi_{z_0} \phi_{z_1} \phi_{z_2}},
\end{equation}
\subsection*{Theoretical effect of model choice on branch length}
\subsection*{Effect of model choice on natural sequences}

\subsubsection*{ExpCM + $\Gamma\omega$ and YNGKP M5}

\subsubsection*{Spielman $\omega_{r}$ values inferred from the ExpCM} 
We inferred the average nonsynonymous fixation rate from the ExpCM following~\citet{spielman2015relationship} as 
\begin{equation}
\label{eq:w_r}
\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times P_{r,xy}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}}
\end{equation}
where $p_{r,x}$ is the stationary state of the ExpCM at site $r$ and codon $x$, $P_{r,xy}$ is the substitution rate from codon $x$ to codon $y$ at site $r$, $Q_{xy}$ is the mutation rate from codon $x$ to codon $y$, and $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.

\subsubsection*{Expected pairwise amino-acid identity}
\textit{Do I need to talk about the branchScale scaling I used?}
The expected pairwise amino-acid identity at a site $r$ over time $t$ for a given model is 
\begin{equation}
\label{eq:f}
\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} [M_{r}\left(t\right)]_{xy}
\end{equation}
where $a$ is all amino acids, $p_{r,x}$ is the stationary state of the model at site $r$ and codon $x$, and $[M_{r}\left(t\right)]_{xy}$ is the transition rate from codon $x$ to codon $y$ at site $r$ given time $t$.

@@

\subsection*{Model Parameters for the simulations}

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_doud}}
\caption{\label{suppfig:prefs_doud}
\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.12$)} 
}
\end{suppfig}

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_lee}}
\caption{\label{suppfig:prefs_lee}
\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.28$)}
 }
\end{suppfig}

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_average}}
\caption{\label{suppfig:prefs_average}
\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by \textit{Lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_average}A  ($\beta = 1.70$)}}
\end{suppfig}

@@

>>>>

We can clearly observe how shifting amino-acid preferences degrade the accuracy of the ExpCM by fitting them to containing increasingly diverged sequences.
Specifically, for H1 and H3 HAs, we created three phylogenetic trees: a ``low'' divergence tree that contains only sequences with $\ge 59\%$ amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with $\ge 46\%$ amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
\ref{fig:compete} shows the subtrees containing each of these sets of HA sequences.
For each subtree, we examined the congruence between site-specific natural selection and the amino-acid preferences measured in the deep mutational scanning experiment. 
We used the ExpCM stringency parameter $\beta$ to measure this congruency~(\ref{eq:Frxy}). 
In essence, $\beta$ rescales the deep mutational scanning measurements~(\ref{fig:compete}). 
Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with a greater stringency, indicating a strong congruence between natural selection and the experimental preferences.

@@

\section*{Materials and Methods}

\subsection*{Models}

\subsubsection*{Experimentally Informed Codon Model (ExpCM)}

The ExpCM is described in detail in \citet{hilton2017phydms}.

In an ExpCM, rate of substitution $P_{r,xy}$ of site $r$ from codon $x$ to $y$ is written in mutation-selection form~\citep{halpern1998evolutionary,mccandlish2014modeling,spielman2015relationship} as
\begin{equation}
P_{r,xy} = Q_{xy} \times F_{r,xy}
\end{equation}
where $Q_{xy}$ is proportional to the rate of mutation from $x$ to $y$, and $F_{r,xy}$ is proportional to the probability that this mutation fixes.
The rate of mutation $Q_{xy}$ is assumed to be uniform across sites, and takes an HKY85-like~\citep{hasegawa1985dating} form.

@@

\subsubsection*{GY94}

We consider the basic Goldman-Yang style (GY94) substitution model defined as M0 in \citet{yang2000codon}. 
$P_{xy}$ is the substitution rate from codon $x$ to codon $y$ and is defined by

\begin{equation}
\label{eq:GY94}
P_xy = 
\begin{cases}
  0 & \mbox{if $x$ and $y$ differ by more than one nucleotide,}\\
  \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transversion,} \\
  \kappa \mu \omega \Phi_{y} & \mbox{if $x$ is converted to $y$ by a single-nucleotide transition,} \\
  -\sum\limits_{z \ne x} P_{xz} & \mbox{if $x = y$.}
  \end{cases}
\end{equation}
where $\kappa$ is the transition-transversion ratio, $\Phi_y$ is the equilibrium frequency of
codon $y$, $\omega$ is the gene-wide rate of non-synonymous change, and $\mu$ is the substitution rate.

\subsubsection*{GY94 with $\omega_r$}
Following~\citet{spielman2015relationship}, we inferred an $\omega$ for each site in the protein ($\omega_r$) from an ExpCM. 
\begin{equation}
\label{eq:w_r}
\omega_{r} = \frac{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times \frac{P_{r,xy}}{\omega}}}{\sum_{x} \sum_{y \in N_x} {p_{r,x} \times Q_{xy}}}
\end{equation}
where $p_{r,x}$ is the stationary state of the ExpCM at site $r$ and codon $x$, $P_{r,xy}$ is the substitution rate from codon $x$ to codon $y$ at site $r$, $Q_{xy}$ is the mutation rate from codon $x$ to codon $y$, and $N_x$ is the set of codons that are nonsynonymous to codon $x$ and differ from codon $x$ by only one nucleotide.

We replaced the $\omega$ value in \ref{eq:GY94} with $\omega \times \omega_r$, resulting in a site-specific model with the GY94 stationary state.

\subsection*{Adding $\Gamma\omega$}

For the GY94+$\Gamma\omega$ and ExpCM+$\Gamma\omega$, $\omega$ is drawn from $K=4$ discrete categories and equal weight is given to each category. 
Set to the mean of the discretized $\Gamma$-distribution.

\subsection*{Deep Mutational Scanning Amino-Acid preferences.}
We used deep mutational scanning amino-acid preferences from an H1 HA~\citep{doud2016accurate} and an H3 HA~\citep{lee2018deep} to define the ExpCM's. 
We only considered homologous sites between the H1 and H3 protein (described in \ref{suppfile:WSN_Perth_map}). 
The resulting preference sets, including the average between the H1 and H3 preferences, are found in \ref{suppfile:H1_prefs}, \ref{suppfile:H3_prefs}, and \ref{suppfile:avg_prefs}.

\subsection*{Influenza virus hemagglutinin (HA) sequences and tree topology.}

We downloaded all full-length, coding sequences for 14 of the 18 HA subtypes, excluding the rare subtypes 12, 15, 17, and 18. 
We filtered and aligned the sequences using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}. 
Specifically, we used \texttt{phydms\_comprehensive} default parameters except \texttt{--minidentity 0.3} and we removed sites which are not homologous between the H1 and H3 deep mutational references sequences using the sequence in \ref{suppfile:hybrid_sequence}. 
We subsampled the sequences so that there were 5 sequences per subtype with $\le 1$ sequence per year per subtype. 
%Finally, we ensured the ``landmark" sequences from the H1, H2, and H3 subtypes (\ref{suppfile:required_seqs}) were included. 
The resulting alignment contains 87 sequences and is found in \ref{suppfile:alignment_high}.

We created four subalignments with ``low" and ``intermediate" divergence from either the H1 or the H3 deep mutational scanning reference sequence. 
The ``low divergence" alignments had   $\ge 59\%$ amino-acid identity from the reference sequence and the ``intermediate divergence" alignments had $\ge 46\%$ identity from the reference sequence.
The resulting alignments are found in \ref{suppfile:alignment_lowH1}, \ref{suppfile:alignment_lowH3}, \ref{suppfile:alignment_intermediateH1}, and \ref{suppfile:alignment_intermediateH3}.

We inferred the tree topology of each alignment using \texttt{RAxML}~\citep{stamatakis2006raxml} and the GTRCAT model. 
We estimated the branch lengths of given this fixed topology for the ExpCM and GY94 models using \texttt{phydms\_comprehensive}~\citep{hilton2017phydms}.

\subsection*{Asymptotic sequence divergence.}

We built ExpCM and GY94 models using parameters fit to the ``low divergence from H1" tree.  
Specifically, we built a GY94 model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94}, a GY94+$\Gamma\omega$ model using the parameter values in \ref{suppfile:hybrid_lowH1_GY94+Gr}, an ExpCM using the parameters in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the amino-acid preferences in \ref{suppfile:H1_prefs}, and a GY94+$\omega_r$ using the GY94 parameters in \ref{suppfile:hybrid_lowH1_GY94} and the ExpCM defined by\ref{suppfile:hybrid_lowH1_ExpCMdoud} and \ref{suppfile:H1_prefs}.

For each model we calculated the expected sequence divergence given some branch length $t$ by 
\begin{equation}
\label{eq:f}
\sum_a \sum_{x \in a} p_{r,x} \sum_{y \in a} [M_{r}\left(t\right)]_{xy}
\end{equation}
where $a$ is all amino acids, $p_{r,x}$ is the stationary state of the model at site $r$ and codon $x$, and $[M_{r}\left(t\right)]_{xy}$ is the transition rate from codon $x$ to codon $y$ at site $r$ given time $t$.

\subsection*{Simulations}

We simulated sequences using \texttt{pyvolve}~\citep{spielman2015pyvolve} along tree for the full alignment (\ref{fig:empirical_trees};\ref{suppfile:alignment_high}) using an ExpCM defined by the parameters in \ref{suppfile:hybrid_lowH1_ExpCMdoud} and the amino-acid preferences in \ref{suppfile:H1_prefs}. 
We performed 10 replicate simulations and estimated the branch lengths for each replicate using \texttt{phydms\_comprehensive}.

@@

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_doud}}
\caption{\label{suppfig:prefs_doud}
\textbf{H1 preferences measured by \cite{doud2016accurate} (\ref{suppfile:H1_prefs}) rescaled with the ExpCM stringency parameter optimized in \ref{fig:empirical_trees}  ($\beta = 1.12$)}. 
}
\end{suppfig}

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_lee}}
\caption{\label{suppfig:prefs_lee}
\textbf{H3 preferences measured by~\citet{lee2018deep} (\ref{suppfile:H1_prefs}) rescaled with the ExpCM stringency parameter optimized in \ref{fig:empirical_trees}  ($\beta = 1.28$)}
 }
\end{suppfig}

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_average}}
\caption{\label{suppfig:prefs_average}
\textbf{The average of the H1 preferences measured by \cite{doud2016accurate} and the H3 preferences measured by~\citet{lee2018deep} ((\ref{suppfile:avg_prefs})) rescaled with the ExpCM stringency parameter optimized in \ref{fig:empirical_trees}  ($\beta = 1.70$)}}
\end{suppfig}

@@

\begin{suppfig}[H]
\centerline{\includegraphics[width=0.50\textwidth]{figures/divergence_distances.pdf}}
\caption{\label{suppfig:subalignments}
\textbf{Subalignments.}
}
\end{suppfig}

\begin{suppfile}
\caption{
\label{suppfile:hybrid_sequence}
The ``hybrid" sequence used to align the HA sequences.
Specifically, this sequence is the H1 deep mutational scanning reference sequence A/Wilson Smith/1933 except it only contains the codon positions shared between A/Wilson Smith/1933 and the H3 deep mutational scanning reference sequence A/Perth/2009. 
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:required_seqs}
These sequences from the H1, H2, and H3 subtypes were required to be in the final alignment. 
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:alignment_high}
HA sequences used in \ref{fig:empirical_trees}.  
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:alignment_lowH1}
HA sequences for the ``low divergence from H1" tree in \ref{fig:compete}.  
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:alignment_lowH3}
HA sequences for the ``low divergence from H3" tree in \ref{fig:compete}.  
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:alignment_intermediateH1}
HA sequences for the ``low divergence from H1" tree in \ref{fig:compete}.  
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:alignment_intermediateH3}
HA sequences for the ``low divergence from H3" tree in \ref{fig:compete}.  
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:WSN_Perth_map}
Homologous sites in WSN and Perth using sequential numbering starting at the second (first non-M) codon. 
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:H1_prefs}
Deep mutational scanning preferences from \citet{doud2016accurate} for the homologous sites between the H1 reference sequence and the H3 reference sequence. 
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:H3_prefs}
Deep mutational scanning preferences from \citet{lee2018deep} for the homologous sites between the H1 reference sequence and the H3 reference sequence. 
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:avg_prefs}
The average deep mutational scanning preferences from \citet{doud2016accurate} ad \citet{lee2018deep}. 
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:hybrid_lowH1_ExpCMdoud}
ExpCM model parameters fit using the amino-acid preferences from~\citet{doud2016accurate}(\ref{suppfile:H1_prefs}) and the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1})
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:hybrid_lowH1_GY94}
GY94 model parameters fit using the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1}). 
}
\end{suppfile}

\begin{suppfile}
\caption{
\label{suppfile:hybrid_lowH1_GY94+Gr}
GY94+$\Gamma\omega$ model parameters fit using the ``low divergence from H1" tree (\ref{suppfile:alignment_lowH1}). 
}
\end{suppfile}
